{\huge Supplements for: Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2024-02-05 Docs/manuscriptRev.Rnw
Edits to address some of the reviewer concerns.

\begin{abstract}
Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions. Yet, we often lack this information entirely or find conflicting evidence across studies. The latter is the case for shifts in spring phenology---one of the most reported and consistent impacts of anthropogenic climate change, and also one of the most critical to forecasting, given its role in carbon sequestration. Some common garden studies have found evidence of important provenance effects, which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence. Here, we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided. Across continents, we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events. These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America. Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events. 
\end{abstract} %Spring tree phenology is critical to forecasting ... Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events. We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data  for  seven angiosperm and eight gymnosperm species from 17 studies, from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.  We found ... This is important because...

\noindent \emph{Keywords: budburst, budset, climate change, common gardens, forests, leafout, provenance, spring phenology}\\

@@

Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies. In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenotypes \citep{AitkenBemmels16, Alberto13}. Such work has established common clines in fall phenology, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset). Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows levels of adaptation that may be critical to forecasting and mitigation \citep{vitasse2009,Basler:2012}.

@@

To test for evidence of adaptation in spring phenology and what factors may underlies differences observed across studies, we comprehensively examined clines for spring events, including fall events when possible. We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, including for differences between: (1) spring and fall phenology, (2) studies in Europe and North America, (3) angiosperm and gymnosperm species. To address these questions, we combined Bayesian hierarchical models with a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology.

@@

\section{Results}
Our final dataset included seven deciduous angiosperm and eight evergreen gymnosperm species from 17 studies and 19 gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species (five in North America and two in Europe) also had fall event information available. Most species in North American gardens were gymosperms (7/11 species) while most species in European gardens were angiosperms (3/4 species).

@@

Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend---albeit a much weaker one than for fall phenology. We found angiosperm versus gymnosperm species diverged in their clines with MAT. Combined with our finding of much stronger clines for fall phenology in North America, these results support the idea of potential variation across continents and/or species type that may underlie the debate in whether spring events show important clinal variation. As clines with spring events were very weak, however, and gardens almost always focus on only one species, understanding these diverging results well enough to aid forecasting would take significant additional investment in common garden studies. %EMWSep4 -- we need to be careful in how many times we say 'we need more data' -- it's a common need, so it's better to focus on specifics of what the data suggests, or doesn't and say we need more data -- at most once in a paper. More on this below.

@@

%EMWSep30: Please double-check we're explaining trends correctly before you submit; I think we are -- but please check closely.
Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events. Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction. Such differences could be driven by their varying investment strategies. As gymnosperms photosynthesize without leafout they generally leafout much later than angiosperms, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk more than angiosperms. In contrast, angiosperms may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}. Testing these hypotheses would require more information on frost risk and forest community assembly from across the provenance locations. 
%EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ... and I am not 100% sure it makes sense -- both angiosperms and conifers need to activate their xylem to photosynthesize I believe. 
% Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

@@

%EMWSep30: Updated figures are nice! I shortened the descriptions in both and re-arranged a little. Try to keep phrasing more consistent here (don't mix 'as we move north' with 'per degree northward' -- pick one). Check my edits closely to make sure they're correct. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
    \caption{Effects of latitude on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text). Zero---no effect---is shown with a dashed line. Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America. Spring events slightly advanced in gymnosperms and delayed in angiosperms per degree northward. Fall events advanced per degree northward for both species types.}
    \label{figure:continental_spptype_lat}
\end{figure}

>>>>

\begin{abstract} %EWMFeb5 -- edited to define provenance differences 
Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions. Yet, we often lack this information entirely or find conflicting evidence across studies. The latter is the case for shifts in spring phenology---one of the most reported and consistent impacts of anthropogenic climate change, and also one of the most critical to forecasting, given its role in carbon sequestration. Some common garden studies have found evidence of important provenance effects (i.e. population differences), which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence. Here, we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided. Across continents, we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events. These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America. Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events. 
\end{abstract} %Spring tree phenology is critical to forecasting ... Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events. We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data  for  seven angiosperm and eight gymnosperm species from 17 studies, from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.  We found ... This is important because...

\noindent \emph{Keywords: budburst, leafout, budset, senescence, climate change, common gardens, deciduous and evergreen trees, provenance, spring phenology}\\ %EWMFeb5 -- if you have an abundance of keywords, remove ones that are redundant with title/abstract

@@

%EWMFeb5 -- edits below to define provenance; and added a reference to 'local adaptation'
Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies. In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins (called `provenances' often in forestry) under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenotypes \citep{AitkenBemmels16, Alberto13}. Such work has established common clines in fall phenology suggestive of local adaptation, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset). Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows levels of adaptation that may be critical to forecasting and mitigation \citep{vitasse2009,Basler:2012}.

@@

To test for evidence of adaptation in spring phenology and what factors may underlies differences observed across studies, we comprehensively examined clines for spring events, including fall events when possible. We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, including for differences between: (1) spring and fall phenology, (2) studies in Europe and North America, (3) angiosperm and gymnosperm species, which represent a deep evolutionary split in the plant tree of life. To address these questions, we combined Bayesian hierarchical models with a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology. %EWMFeb5 -- edits above to deal with angio/gymno

@@

\section{Results}
%EWMFeb5 -- we do a nice job here of mentioning the angiosperms are deciduous and the gymnosperms are evergreen. Could highlight this in response letter. 
Our final dataset included seven deciduous angiosperm and eight evergreen gymnosperm species from 17 studies and 19 gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species (five in North America and two in Europe) also had fall event information available. Most species in North American gardens were gymosperms (7/11 species) while most species in European gardens were angiosperms (3/4 species).

@@

%EWMFeb5 -- small edits below about angio/gymno
Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend---albeit a much weaker one than for fall phenology. We found angiosperm (all deciduous) versus gymnosperm (all evergreen) species diverged in their clines with MAT. Combined with our finding of much stronger clines for fall phenology in North America, these results support the idea of potential variation across continents and/or species type that may underlie the debate in whether spring events show important clinal variation. As clines with spring events were very weak, however, and gardens almost always focus on only one species, understanding these diverging results well enough to aid forecasting would take significant additional investment in common garden studies. %EMWSep4 -- we need to be careful in how many times we say 'we need more data' -- it's a common need, so it's better to focus on specifics of what the data suggests, or doesn't and say we need more data -- at most once in a paper. More on this below.

@@

%EMWSep30: Please double-check we're explaining trends correctly before you submit; I think we are -- but please check closely.
%EWMFeb5 -- more edits here about angio/gymno
Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events. Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction. Such differences could be driven by the varying investment strategies, given that all our studied gymnosperms were evergreen and all angiosperms were deciduous. As evergreen species photosynthesize without leafout they generally leafout much later than deciduous species, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk. In contrast, deciduous species may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}. Testing these hypotheses would require more information on frost risk and forest community assembly from across the provenance locations, but seems an important step towards understanding the drivers of this variation. Without a clear mechanism, extrapolating these results to other species or across Europe and North America may be difficult, especially given biases in the data. %EWMFeb5 -- trying to address request by R2 about extrapolating while also keeping ms flow. 
%EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ... and I am not 100% sure it makes sense -- both angiosperms and conifers need to activate their xylem to photosynthesize I believe. 
% Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

@@

%EMWSep30: Updated figures are nice! I shortened the descriptions in both and re-arranged a little. Try to keep phrasing more consistent here (don't mix 'as we move north' with 'per degree northward' -- pick one). Check my edits closely to make sure they're correct. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
    \caption{Effects of latitude on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text). Zero---no effect---is shown with a dashed line. Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America. Spring events slightly advanced in gymnosperms (all evergreen species) and delayed in angiosperms (all deciduous species) per degree northward. Fall events advanced per degree northward for both species types.} %EWMFeb5 -- more edits for angio/gymno
    \label{figure:continental_spptype_lat}
\end{figure}
